OpenMPIR

### 2017-09-07 paper.tex
Added citations

Compilers use internal representations (IRs) for implementing analyses and
optimizations. Historically, a major drawback for parallel programs is that IRs
haven't had the ability to represent, and therefore reason about, parallelism.
This has prevented the application of common optimizations to concurrent code.
The recent work of Schardl et al.  \cite{shardl2017} has shown that real world
compiler IRs can be extended to represent fork-join parallelism. By extending
the LLVM instruction set with three instructions, they were able to capture all
of the semantics to implement the semantics of Cilk, an extension to C. They
suggested that a similar approach could be taken with OpenMP tasks.

@@

The remainder of the paper is organized as follows: In Section~\ref{Sec:background} 
we give background and motivation for the paper. Following that, we describe OpenMP
tasking in Section~\ref{Sec:OpenMP}, and Tapir in Section~\ref{Sec:Tapir}. We then 
discuss the implementation, and how we compile OpenMP task constructs to Tapir IR in
Section~\ref{Sec:Implementation}. In Section~\ref{Sec:Evaluation} we describe the 
evaluation setup, including what implementations we compare to and how. We discuss
the results of evaluation in Section~\ref{Sec:Results}, including possible
explanations for discrepancies. We discuss the implications of the results ans the
significance of this work in a large context in~\ref{Sec:Discussion}. Finally,
we cover some of the many ways the work can be extended in
Section~\ref{Sec:Future}, and conclude in Section~\ref{Sec:Conclusion}.

@@

\wmnote{This feels like we're trying to convince someone about LLVM: perhaps
LLVM has gained traction in the academic community for both its simplicity and
relatively painless modification} 
\gsnote{I wanted to give a brief nod to the importance of LLVM, given the venue
(LLVM in HPC). Is this any more satisfactory?} 
Among IRs, LLVM is arguably the most ubiquitous \cite{llvm}. Winner of the
2012 ACM Software System Award, it is in use by frontends ranging from Haskell
to C++, and backends ranging from x86 to PTX \cite{}. Implementing a simple,
extensible single static assignment IR, LLVM enables powerful optimizations
that are effective for a wide range of programming languages and hardware
\cite{llvm}.

@@

In this section we further discuss a significant challenge in HPC:
fragmentation of parallel programming models. There are many, many, many ways
to write parallel programs \cite{....}. From libraries, to language extensions,
to standalone languages, to combinations of the above, fragmentation of
parallel programming models is a problem that is only getting worse. This means
that when writing tooling, optimizations, or analyses, one has to choose which
model to target. One common difficulty among all of these is that reasoning
about parallel programs is hard \cite{}, and compilers can help \cite{}. This
is exactly analogous the problem LLVM solves: compiler optimizations are hard
to do well, so sharing them is valuable. Tapir is a demonstration that the 
the solutions in the case of serial code can be effectively extended to the 
case of parallel code.

@@

\section{Implementation}

@@

For evaluation we compared performance on the Barcelona OpenMP task suite to 
existing OpenMP implementations. The Barcelona OpenMP task suite is a set of 
tests intended to test performance of OpenMP implementations using both
irregular and regular tasking. The benchmarks are described in
Figure~\ref{Fig:benchmark_desc}. The benchmark suite has added an unbalanced
tree search benchmark since the paper was published.  Unfortunately, this seems
to be buggy, and failed on every implementation for even medium input sizes. We
have therefore left it out of our evaluation.

@@

file  against  every  other  sequence  using  the Myers  and Miller \cite{}

@@

of a vector of n complex values using the Cooley-Tukey \cite{cooley-tukey}

@@

\item \texttt{Health}: simulates de Columbian Health Care System \cite{?}. It

@@

for  each  village being  simulated.  Once  the lower   levels   have   been
simulated synchronization   occurs. 

@@

fast  parallel  sorting  variation  \cite{mergesort}  of  the  ordinary
mergesort.  First, it divides an array of elements in two halves, sorting  each
half  recursively,  and  then  merging  the  sorted halves with a parallel
divide-and-conquer method rather than the  conventional  serial  merge.  Tasks
are  used  for  each  split and merge. When the array is too small, a serial
quicksort is used  so increase  the  task  granularity.  To  avoid  the
overhead of  quicksort,  an insertion  sort  is  used  for  very  small  arrays
(below a threshold of 20 elements).

@@

solution with tasks parallelism seems to obtain better results \cite{}. In each

@@

matrix for multiplication of large dense matrices \cite{}. Decomposition is
done by dividing each dimension of the matrix into  two  sections  of  equal
size. For each decomposition a task is created. 

>>>>

Compilers use internal representations (IRs) for implementing analyses and
optimizations. Historically, a major drawback for parallel programs is that IRs
haven't had the ability to represent, and therefore reason about, parallelism.
This has prevented the application of common optimizations to concurrent code.
The recent work of Schardl et al. \cite{tapir} has shown that real world
compiler IRs can be extended to represent fork-join parallelism. By extending
the LLVM instruction set with three instructions, they were able to capture all
of the semantics to implement the semantics of Cilk, an extension to C. They
suggested that a similar approach could be taken with OpenMP tasks.

@@

The remainder of the paper is organized as follows: In Section~\ref{Sec:Background} 
we give background and motivation for the paper. Following that, we describe OpenMP
tasking in Section~\ref{Sec:OpenMP}, and Tapir in Section~\ref{Sec:Tapir}. We then 
discuss the implementation, and how we compile OpenMP task constructs to Tapir IR in
Section~\ref{Sec:Implementation}. In Section~\ref{Sec:Evaluation} we describe the 
evaluation setup, including what implementations we compare to and how. We discuss
the results of evaluation in Section~\ref{Sec:Results}, including possible
explanations for discrepancies. We discuss the implications of the results ans the
significance of this work in a large context in~\ref{Sec:Discussion}. Finally,
we cover some of the many ways the work can be extended in
Section~\ref{Sec:Future}, and conclude in Section~\ref{Sec:Conclusion}.

@@

\wmnote{This feels like we're trying to convince someone about LLVM: perhaps
LLVM has gained traction in the academic community for both its simplicity and
relatively painless modification} 
\gsnote{I wanted to give a brief nod to the importance of LLVM, given the venue
(LLVM in HPC). Is this any more satisfactory?} 
Among IRs, LLVM is arguably the most ubiquitous \cite{lattner2004llvm}. Winner
of the 2012 ACM Software System Award, it is in use by frontends ranging from
Haskell to C++, and backends ranging from x86 to PTX \cite{lattner2004llvm}.
Implementing a simple, extensible single static assignment IR, LLVM enables
powerful optimizations that are effective for a wide range of programming
languages and hardware \cite{lattner2004llvm}.

@@

In this section we further discuss a significant challenge in HPC:
fragmentation of parallel programming models. There are many ways to write
parallel programs \cite{spec, qthreads, chapel, cilk, kokkos, legion}. From
libraries, to language extensions, to standalone languages, to combinations of
the above, fragmentation of parallel programming models is a problem that is
only getting worse. This means that when writing tooling, optimizations, or
analyses, one has to choose which model to target. One common difficulty among
all of these is that reasoning about parallel programs is hard, and compilers
can help. This is exactly analogous the problem LLVM solves: compiler
optimizations are hard , so sharing them is valuable. Tapir is a demonstration
that the the LLVM approach to compilation of serial code can be effectively
extended to the case of parallel code.

@@

\section{Implementation} \label{Sec:Implementation}

@@

For evaluation we compared performance on the Barcelona OpenMP task suite
\cite{barcelona} to existing OpenMP implementations. The Barcelona OpenMP task
suite is a set of tests intended to test performance of OpenMP implementations
using both irregular and regular tasking. The benchmark suite has added an
unbalanced tree search benchmark since the paper was published. Unfortunately
we were unable to succesfully run it on any implementation, even for even
medium input sizes. We have therefore left it out of our evaluation. The
following descriptions are abridged versions taken verbatim from
\cite{barcelona}:

@@

file  against  every  other  sequence  using  the Myers and Miller 

@@

of a vector of n complex values using the Cooley-Tukey 

@@

\item \texttt{Health}: simulates the Columbian Health Care System. It

@@

for  each  village being  simulated. Once the lower levels have been simulated
synchronization occurs. 

@@

fast  parallel  sorting  variation of  the  ordinary mergesort.  First, it
divides an array of elements in two halves, sorting  each half  recursively,
and  then  merging  the  sorted halves with a parallel divide-and-conquer
method rather than the  conventional  serial  merge.  Tasks are  used  for
each  split and merge. When the array is too small, a serial quicksort is used
so increase  the  task  granularity.  To  avoid  the overhead of  quicksort, an
insertion  sort  is  used  for  very  small  arrays (below a threshold of 20
elements).

@@

solution with tasks parallelism seems to obtain better results. In each

@@

matrix for multiplication of large dense matrices. Decomposition is done by
dividing each dimension of the matrix into  two  sections  of  equal size. For
each decomposition a task is created. 
